---
title: "Terns mass trends over time"
subtitle: "By age class and breeding status - issue #5 task 3"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.width = 8, fig.height = 5)
library(readxl)
library(dplyr)
library(lubridate)
library(tidyr)
library(ggplot2)
library(knitr)
```

# Background

Issue #5 task 3 asks whether there's a temporal trend in body condition
(mass vs. year), per species. Extended per Inbal's request to compare that
trend across **three groups**, per species:

1. **Juveniles** - this-year fledglings (EURING age 3), representing
   post-fledging condition.
2. **Adult breeders** - spring-caught adults confirmed to have stayed to
   breed (the task 2c / any-year site-fidelity definition from
   `02_mass_analysis.Rmd`).
3. **Adult non-breeders** - spring-caught adults with no summer sighting in
   any year (presumed passage migrants, not confirmed as Atlit breeders).

This checks whether a mass trend (if any) is driven by a specific
age/breeding-status class, rather than the population as a whole.

# Data loading

```{r load-data}
raw_dir <- "../../data_raw"
if (!dir.exists(raw_dir)) raw_dir <- "../data_raw"
if (!dir.exists(raw_dir)) raw_dir <- "data_raw"
stopifnot(dir.exists(raw_dir))

proc_dir <- "../../data_processed"
if (!dir.exists(proc_dir)) proc_dir <- "../data_processed"
if (!dir.exists(proc_dir)) proc_dir <- "data_processed"

out_dir <- "../output"
if (!dir.exists(out_dir)) out_dir <- "output"

ringing_path <- file.path(raw_dir, "ringing_data_raw.xlsx")
stopifnot(file.exists(ringing_path))

col_names_47 <- c(
  "CR1","RingReplace","Rec","RingPrefix","RingNum","Species","Sex","Age","Wing","Weight",
  "Date","Time","Place","Lat","Lon","Ringer","Country","Remarks","Net","CR2","Subsp","Tail","Head",
  "Breeding","CP1","CP2","CP3","CP4","Colour","D","M","OrigDate","OrigCountry","OrigLat","OrigLon","OrigAge",
  "OldCR1","OldCR2","OldCR3","OldMetal1","OldMetal2","OilHead","OilWings","OilUpperP","OilUnderP","OilLegs","OilSum"
)
raw <- read_excel(ringing_path, sheet = "Data", skip = 2, col_names = col_names_47)

terns <- raw %>%
  filter(Species %in% c("STEALB", "STEHIR")) %>%
  mutate(
    year = year(Date),
    month = month(Date),
    ring_id = paste0(trimws(RingPrefix), RingNum),
    is_new = is.na(Rec) | trimws(Rec) == "",
    species_label = ifelse(Species == "STEHIR", "Common Tern", "Little Tern")
  )
```

# Building the three comparison groups

## Juveniles (age 3, Jun-Aug)

Unlike the adult analyses (spring, Mar-May), juveniles can only be
measured once they're in the nest/newly fledged, i.e. during the breeding
season. Window is Jun-**Aug** here (wider than the Jun-Jul used for adult
breeding-season mass in task 2) because juvenile catches remain
substantial into August (e.g. Common Tern age-3 catches: Jun=106,
Jul=1442, Aug=1049 - dropping the August tail would discard a large,
non-trivial share of the juvenile data for no analogous reason to why
adults were restricted to Jun-Jul).

```{r juvenile-mass}
juvenile_records <- terns %>%
  filter(is_new, Age == 3, !is.na(Weight), month %in% 6:8)

juvenile_mass <- juvenile_records %>%
  group_by(year, Species, species_label) %>%
  summarise(mean_weight = mean(Weight), sd_weight = sd(Weight), n = n(), .groups = "drop") %>%
  mutate(group = "Juvenile")

kable(juvenile_mass %>% select(-species_label), digits = 2, caption = "Juvenile (age 3) mean weight, Jun-Aug")
```

## Adult breeders vs. non-breeders (spring weight, any-year confirmed)

Same construction as task 2c in `02_mass_analysis.Rmd`: any adult (new
ringing or recapture) weighed in spring (Mar-May), flagged as a confirmed
breeder if that ring was also recorded (recapture or resighting) in
June/July of **any** year.

```{r adult-breeder-mass}
juvenile_ages <- c(1, 3)

summer_ring_ids_anyyear <- terns %>%
  filter(month %in% 6:7) %>%
  distinct(ring_id) %>%
  pull(ring_id)

spring_adults_anyyear <- terns %>%
  filter(!(Age %in% juvenile_ages), !is.na(Weight), month %in% 3:5) %>%
  mutate(confirmed_breeder_anyyear = ring_id %in% summer_ring_ids_anyyear)

adult_mass_by_year <- spring_adults_anyyear %>%
  group_by(year, Species, species_label, confirmed_breeder_anyyear) %>%
  summarise(mean_weight = mean(Weight), sd_weight = sd(Weight), n = n(), .groups = "drop") %>%
  mutate(group = ifelse(confirmed_breeder_anyyear, "Adult breeder", "Adult non-breeder")) %>%
  select(-confirmed_breeder_anyyear)

kable(adult_mass_by_year %>% select(-species_label), digits = 2, caption = "Adult (spring) mean weight, breeders vs. non-breeders")
```

## All adults (unsplit)

The plain version of task 3's original ask - all spring-caught adults
together, no breeder/non-breeder split. Kept as a reference point: this is
just breeders + non-breeders combined, so it's shown alongside the split
groups in the plot and the trend-statistics table below, but it's
deliberately **excluded from the interaction test** (it's not a
mutually-exclusive category vs. the other three - including it there would
just be comparing a group against its own subsets).

```{r adult-all-mass}
all_adults_mass <- spring_adults_anyyear %>%
  group_by(year, Species, species_label) %>%
  summarise(mean_weight = mean(Weight), sd_weight = sd(Weight), n = n(), .groups = "drop") %>%
  mutate(group = "All adults (unsplit)")

kable(all_adults_mass %>% select(-species_label), digits = 2, caption = "Adult (spring) mean weight, breeders + non-breeders combined")
```

## Combined

```{r combine-groups}
combined_by_year <- bind_rows(
  juvenile_mass,
  adult_mass_by_year,
  all_adults_mass
) %>%
  mutate(group = factor(group, levels = c("Juvenile", "Adult non-breeder", "Adult breeder", "All adults (unsplit)")))

# Individual-level records, for the regression/trend tests below.
# "All adults (unsplit)" duplicates rows already in Adult breeder/non-breeder
# by construction - fine for the per-group trend_stats table, but must be
# excluded from the interaction test (see note above).
combined_records <- bind_rows(
  juvenile_records %>% mutate(group = "Juvenile"),
  spring_adults_anyyear %>% mutate(group = ifelse(confirmed_breeder_anyyear, "Adult breeder", "Adult non-breeder")),
  spring_adults_anyyear %>% mutate(group = "All adults (unsplit)")
) %>%
  mutate(group = factor(group, levels = c("Juvenile", "Adult non-breeder", "Adult breeder", "All adults (unsplit)")))
```

# Visual comparison

One fully separate panel per species x group combination (8 total) -
Common Tern and Little Tern were sharing a y-axis in the same panel
before, which compressed Little Tern's much smaller/narrower weight range
(~40-50g vs. Common Tern's ~110-140g) into a sliver at the bottom of each
plot. Each panel now gets its own independent y-scale. Point size is
proportional to that year's sample size `n` (small-n years, e.g. Little
Tern juveniles in the most recent seasons, are visually obvious rather
than looking like equally-weighted data points).

**Trend line fix**: the smooth/trend line is now fit on the same
individual-level data as the `trend_stats` table below, NOT on the
per-year mean points shown in the plot. Fitting on the yearly means (as
before) treats every year equally regardless of how many individuals it's
based on - a thin year (e.g. Little Tern all-adults 2024, n=1) gets the
same say as a year with 80+ individuals. That mismatch was visible for
Little Tern all-adults: the table's individual-level slope is -0.16
g/year, but a couple of small-n, high-mean recent years (2024 n=1,
2026 n=10) were pulling the year-mean-based line to apparently *positive*
in the plot. Fitting the smooth on individual records instead makes the
line match the table:

```{r trends-plot, fig.height=13}
panel_levels <- as.vector(t(outer(levels(combined_by_year$group), c("Common Tern", "Little Tern"), paste, sep = " - ")))

# Local copies with a plotting-only panel_label column - deliberately not
# reassigned back to combined_by_year/combined_records, so this doesn't
# leak into the CSVs written out in # Save outputs below.
combined_by_year_panel <- combined_by_year %>%
  mutate(panel_label = factor(paste(group, species_label, sep = " - "), levels = panel_levels))

combined_records_panel <- combined_records %>%
  mutate(panel_label = factor(paste(group, species_label, sep = " - "), levels = panel_levels))

ggplot(combined_by_year_panel, aes(year, mean_weight, color = species_label)) +
  geom_line(alpha = 0.4, linewidth = 0.5) +
  geom_point(aes(size = n), alpha = 0.7) +
  geom_smooth(data = combined_records_panel, aes(x = year, y = Weight, color = species_label),
              method = "lm", se = FALSE, linewidth = 0.8) +
  facet_wrap(~panel_label, ncol = 2, scales = "free_y") +
  scale_size_continuous(range = c(1, 6), name = "n") +
  guides(color = "none") +
  labs(title = "Mean mass by year, age class and breeding status",
       y = "Mean weight (g)", x = "Year") +
  theme_minimal()
```

# Trend statistics

Individual-level linear regression of `Weight ~ year`, per species and
group (not just correlating the per-year means, so each year is weighted
by its actual sample size):

```{r trend-stats}
trend_stats <- combined_records %>%
  group_by(species_label, group) %>%
  filter(n() >= 3) %>%   # need at least 3 years' worth of individual points to fit a trend
  group_modify(~ {
    m <- lm(Weight ~ year, data = .x)
    s <- summary(m)
    tibble(
      n = nrow(.x),
      n_years = n_distinct(.x$year),
      slope_g_per_year = round(coef(m)["year"], 3),
      se = round(s$coefficients["year", "Std. Error"], 3),
      p_value = signif(s$coefficients["year", "Pr(>|t|)"], 3),
      r_squared = round(s$r.squared, 3)
    )
  }) %>%
  ungroup()

# p < 0.001 shown in scientific notation (still 3 sig figs); p >= 0.001
# shown as a plain decimal - kable's default column-wide formatting was
# rendering EVERY value in the column in scientific notation (e.g.
# "1.56e-01") just because the smallest value in the column needed it.
format_pval <- function(p) {
  ifelse(p < 0.001, formatC(p, format = "e", digits = 2), formatC(p, format = "f", digits = 3))
}

trend_stats_display <- trend_stats %>%
  mutate(significant = p_value < 0.05, p_value = format_pval(p_value)) %>%
  mutate(across(-significant, ~ ifelse(significant, paste0("<strong>", .x, "</strong>"), as.character(.x)))) %>%
  select(-significant)

kable(trend_stats_display, escape = FALSE,
      caption = "Weight ~ year regression, by species and group (significant rows, p&lt;0.05, in bold)")
```

# Does the trend differ between groups?

Interaction test (`Weight ~ year * group`) per species - tests whether the
year-slope itself differs significantly across juveniles / non-breeders /
breeders, rather than just looking at each group's trend in isolation:

```{r trend-interaction, results = "asis"}
for (sp in c("Common Tern", "Little Tern")) {
  d <- combined_records %>%
    filter(species_label == sp, group != "All adults (unsplit)") %>%
    droplevels()
  m <- lm(Weight ~ year * group, data = d)
  a <- anova(m)
  p_interaction <- signif(a["year:group", "Pr(>F)"], 3)
  cat("**", sp, "**: year x group interaction p =", p_interaction, "\n\n", sep = "")
}
```

## Interpretation

A significant interaction means the mass trend over time is not the same
across groups (e.g. breeders declining while juveniles don't, or vice
versa) - that's the more informative result for the paper than any single
group's trend alone. A non-significant interaction means whatever trend
exists (if any) is roughly parallel across groups. Read this alongside the
per-group `trend_stats` table above and the plot - small per-year-per-group
sample sizes (see `n` columns) mean isolated significant/non-significant
results for any one group should be treated cautiously.

# Save outputs

```{r save-outputs}
write.csv(combined_by_year %>% select(-species_label), file.path(proc_dir, "mass_trends_by_group.csv"), row.names = FALSE)
write.csv(trend_stats, file.path(proc_dir, "mass_trends_stats.csv"), row.names = FALSE)
cat("Wrote data_processed/mass_trends_by_group.csv and mass_trends_stats.csv\n")
```
