---
title: "Terns wing-normalized mass trends over time"
subtitle: "Issue #6 - follow-up to issue #5 task 3, per Yosef's suggestion"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.width = 8, fig.height = 5)
library(readxl)
library(dplyr)
library(lubridate)
library(tidyr)
library(ggplot2)
library(knitr)
```

# Background

Issue #5 task 3 found that confirmed breeders show a significantly steeper
mass decline over time than juveniles or non-breeders, in both species (see
`03_mass_trends.Rmd`). Yosef suggested normalizing mass to wing (chord)
length before drawing conclusions from a mass trend, on the theory that raw
mass could partly reflect body-size differences in the sample rather than
body condition per se.

This is not a settled methods question. A simple mass/wing ratio assumes a
proportional, through-origin relationship between the two, which body mass
rarely follows exactly. The better-supported alternative in the literature is
the **scaled mass index (SMI)** (Peig & Green 2009): fit a standardized major
axis (SMA) regression of `log(Weight)` on `log(Wing)`, then rescale each
individual's mass to what it would be at the group's mean wing length, using
the SMA slope as the scaling exponent. Both are computed here, side by side,
rather than picking one upfront.

Same groups as issue #5 task 3: **Juvenile** (age 3, Jun-Aug) plus spring
adults, split into **Adult breeder** / **Adult non-breeder** (task 2c
any-year confirmed-breeder split) **for Common Tern only**. **Little Tern
adults are kept as a single unsplit `Adult` group** - #5 task 2 found no
significant mass difference between Little Tern breeders and non-breeders
under either definition, so splitting an already-small sample here adds
nothing.

# Data loading

```{r load-data}
raw_dir <- "../../data_raw"
if (!dir.exists(raw_dir)) raw_dir <- "../data_raw"
if (!dir.exists(raw_dir)) raw_dir <- "data_raw"
stopifnot(dir.exists(raw_dir))

proc_dir <- "../../data_processed"
if (!dir.exists(proc_dir)) proc_dir <- "../data_processed"
if (!dir.exists(proc_dir)) proc_dir <- "data_processed"

out_dir <- "../output"
if (!dir.exists(out_dir)) out_dir <- "output"

ringing_path <- file.path(raw_dir, "ringing_data_raw.xlsx")
stopifnot(file.exists(ringing_path))

col_names_47 <- c(
  "CR1","RingReplace","Rec","RingPrefix","RingNum","Species","Sex","Age","Wing","Weight",
  "Date","Time","Place","Lat","Lon","Ringer","Country","Remarks","Net","CR2","Subsp","Tail","Head",
  "Breeding","CP1","CP2","CP3","CP4","Colour","D","M","OrigDate","OrigCountry","OrigLat","OrigLon","OrigAge",
  "OldCR1","OldCR2","OldCR3","OldMetal1","OldMetal2","OilHead","OilWings","OilUpperP","OilUnderP","OilLegs","OilSum"
)
raw <- read_excel(ringing_path, sheet = "Data", skip = 2, col_names = col_names_47)

terns <- raw %>%
  filter(Species %in% c("STEALB", "STEHIR")) %>%
  mutate(
    year = year(Date),
    month = month(Date),
    ring_id = paste0(trimws(RingPrefix), RingNum),
    is_new = is.na(Rec) | trimws(Rec) == "",
    species_label = ifelse(Species == "STEHIR", "Common Tern", "Little Tern")
  )
```

# Building the three comparison groups

Identical construction to `03_mass_trends.Rmd`, with the added requirement
that `Wing` (not just `Weight`) is present - `Wing`/`Weight` are both
capture-only fields, so this drops very few extra records (checked below).

```{r groups}
juvenile_records <- terns %>%
  filter(is_new, Age == 3, !is.na(Weight), month %in% 6:8) %>%
  mutate(group = "Juvenile")

juvenile_ages <- c(1, 3)
summer_ring_ids_anyyear <- terns %>%
  filter(month %in% 6:7) %>%
  distinct(ring_id) %>%
  pull(ring_id)

spring_adults_anyyear <- terns %>%
  filter(!(Age %in% juvenile_ages), !is.na(Weight), month %in% 3:5) %>%
  mutate(confirmed_breeder_anyyear = ring_id %in% summer_ring_ids_anyyear,
         group = case_when(
           species_label == "Little Tern" ~ "Adult",
           confirmed_breeder_anyyear ~ "Adult breeder",
           TRUE ~ "Adult non-breeder"
         ))

combined_records_raw <- bind_rows(
  juvenile_records,
  spring_adults_anyyear %>% select(-confirmed_breeder_anyyear)
) %>%
  mutate(group = factor(group, levels = c("Juvenile", "Adult", "Adult non-breeder", "Adult breeder")))

n_before <- nrow(combined_records_raw)
combined_records <- combined_records_raw %>% filter(!is.na(Wing))
n_after <- nrow(combined_records)

cat("Records with Weight:", n_before, "- also with Wing:", n_after,
    sprintf("(%.1f%% retained)\n", 100 * n_after / n_before))
```

```{r coverage-table}
kable(
  combined_records %>% group_by(species_label, group) %>%
    summarise(n = n(), n_years = n_distinct(year),
              min_year = min(year), max_year = max(year), .groups = "drop"),
  caption = "Weight+Wing coverage by species and group"
)
```

# Does wing length itself trend over time?

Before normalizing, check whether `Wing` shows its own temporal trend within
each group - if it doesn't, wing-normalized mass should track raw mass
closely, and this adjustment won't tell us much new for that group.

```{r wing-trend}
wing_trend <- combined_records %>%
  group_by(species_label, group) %>%
  filter(n() >= 10, n_distinct(year) >= 3) %>%
  group_modify(~ {
    m <- lm(Wing ~ year, data = .x)
    s <- summary(m)
    tibble(
      n = nrow(.x),
      slope_mm_per_year = round(coef(m)["year"], 4),
      p_value = signif(s$coefficients["year", "Pr(>|t|)"], 3)
    )
  }) %>%
  ungroup()

kable(wing_trend, caption = "Wing ~ year regression, by species and group")
```

# Two normalization methods

## Simple ratio

```{r ratio-mass}
combined_records <- combined_records %>%
  mutate(mass_ratio = Weight / Wing)
```

## Scaled mass index (SMI, Peig & Green 2009)

SMA slope `b` fit **per species x group** (juveniles, breeders and
non-breeders may have genuinely different mass-wing scaling relationships,
so pooling would misrepresent all three) on `log(Weight) ~ log(Wing)`. Each
individual's mass is then rescaled to what it would weigh at that group's
mean wing length:

`SMI_i = Weight_i * (Wing_mean / Wing_i) ^ b`

where `b` is the SMA slope, `sd(log Weight) / sd(log Wing)` (equivalent to
the OLS slope divided by the correlation coefficient).

```{r smi-mass}
compute_smi <- function(df) {
  lx <- log(df$Wing)
  ly <- log(df$Weight)
  b <- sd(ly) / sd(lx)  # SMA slope
  wing_mean <- mean(df$Wing)
  df$smi <- df$Weight * (wing_mean / df$Wing) ^ b
  df$smi_b <- b
  df
}

combined_records <- combined_records %>%
  group_by(species_label, group) %>%
  group_modify(~ compute_smi(.x)) %>%
  ungroup()

kable(
  combined_records %>% distinct(species_label, group, smi_b) %>% arrange(species_label, group),
  digits = 3, caption = "SMA slope (b) used for SMI, by species and group"
)
```

# Trend statistics, all three metrics side by side

```{r trend-stats}
format_pval <- function(p) {
  ifelse(p < 0.001, formatC(p, format = "e", digits = 2), formatC(p, format = "f", digits = 3))
}

fit_trend <- function(df, response) {
  df %>%
    group_by(species_label, group) %>%
    filter(n() >= 3) %>%
    group_modify(~ {
      m <- lm(reformulate("year", response = response), data = .x)
      s <- summary(m)
      tibble(
        n = nrow(.x),
        n_years = n_distinct(.x$year),
        slope_per_year = round(coef(m)["year"], 4),
        se = round(s$coefficients["year", "Std. Error"], 4),
        p_value = signif(s$coefficients["year", "Pr(>|t|)"], 3),
        r_squared = round(s$r.squared, 3)
      )
    }) %>%
    ungroup() %>%
    mutate(metric = response)
}

trend_stats <- bind_rows(
  fit_trend(combined_records, "Weight"),
  fit_trend(combined_records, "mass_ratio"),
  fit_trend(combined_records, "smi")
) %>%
  mutate(metric = recode(metric, Weight = "Raw mass (g)",
                          mass_ratio = "Mass/Wing ratio",
                          smi = "SMI (g, rescaled)"),
         metric = factor(metric, levels = c("Raw mass (g)", "Mass/Wing ratio", "SMI (g, rescaled)"))) %>%
  arrange(species_label, group, metric)

trend_stats_display <- trend_stats %>%
  mutate(significant = p_value < 0.05, p_value_fmt = format_pval(p_value)) %>%
  mutate(p_value_fmt = ifelse(significant, paste0("<strong>", p_value_fmt, "</strong>"), p_value_fmt)) %>%
  select(species_label, group, metric, n, n_years, slope_per_year, se, p_value = p_value_fmt, r_squared)

kable(trend_stats_display, escape = FALSE,
      caption = "value ~ year regression, by species/group/metric (significant rows, p&lt;0.05, in bold)")
```

# Visual comparison

Trend lines are drawn only where that species/group/metric's year slope is
significant (p<0.05, per the `trend_stats` table above) - a flat/absent line
means no detectable trend, rather than implying a weak trend still worth
drawing.

```{r trends-plot, fig.height=13}
long_records <- combined_records %>%
  select(species_label, group, year, Weight, mass_ratio, smi) %>%
  pivot_longer(cols = c(Weight, mass_ratio, smi), names_to = "metric", values_to = "value") %>%
  mutate(metric = recode(metric, Weight = "Raw mass (g)",
                          mass_ratio = "Mass/Wing ratio",
                          smi = "SMI (g, rescaled)"),
         metric = factor(metric, levels = c("Raw mass (g)", "Mass/Wing ratio", "SMI (g, rescaled)")),
         panel_label = paste(group, species_label, sep = " - "))

sig_combos <- trend_stats %>%
  filter(p_value < 0.05) %>%
  distinct(species_label, group, metric)

long_records_sig <- long_records %>% semi_join(sig_combos, by = c("species_label", "group", "metric"))

ggplot(long_records, aes(year, value, color = species_label)) +
  geom_point(alpha = 0.15, size = 0.8) +
  geom_smooth(data = long_records_sig, method = "lm", se = FALSE, linewidth = 0.8) +
  facet_grid(panel_label ~ metric, scales = "free_y") +
  guides(color = "none") +
  labs(title = "Mass trend by year, metric, age class and breeding status",
       subtitle = "Trend line shown only where the year slope is significant (p<0.05)",
       y = NULL, x = "Year") +
  theme_minimal() +
  theme(strip.text.y = element_text(angle = 0, size = 7), strip.text.x = element_text(size = 9))
```

# Does the trend differ between groups? (interaction test, per metric)

```{r trend-interaction, results = "asis"}
for (metric_col in c("Weight", "mass_ratio", "smi")) {
  metric_label <- recode(metric_col, Weight = "Raw mass", mass_ratio = "Mass/Wing ratio", smi = "SMI")
  cat("**", metric_label, "**\n\n", sep = "")
  for (sp in c("Common Tern", "Little Tern")) {
    d <- combined_records %>% filter(species_label == sp) %>% droplevels()
    m <- lm(reformulate(c("year", "group", "year:group"), response = metric_col), data = d)
    a <- anova(m)
    p_interaction <- signif(a["year:group", "Pr(>F)"], 3)
    cat("- ", sp, ": year x group interaction p = ", p_interaction, "\n", sep = "")
  }
  cat("\n")
}
```

## Interpretation

Compare each metric's breeder-vs-non-breeder-vs-juvenile pattern against the
raw-mass result from issue #5 task 3 (Common Tern breeder slope -0.55
g/year, p=1.6e-06; Little Tern breeder slope -0.24 g/year, p=0.007, both
significantly steeper than the other two groups). If the wing-normalized
metrics show the same qualitative pattern (breeders declining faster,
comparable significance), that's reassurance the raw-mass finding isn't an
artifact of changing body size in the sample. If normalization
substantially weakens or removes the breeder-specific decline, that
suggests the raw-mass trend was at least partly a size effect rather than a
condition effect - see the `Wing ~ year` check above for whether that's
plausible per group.

# Save outputs

```{r save-outputs}
write.csv(combined_records %>% select(-smi_b),
          file.path(proc_dir, "normalized_mass_records.csv"), row.names = FALSE)
write.csv(trend_stats, file.path(proc_dir, "normalized_mass_trends_stats.csv"), row.names = FALSE)
cat("Wrote data_processed/normalized_mass_records.csv and normalized_mass_trends_stats.csv\n")
```
