---
title: "Terns body mass analysis"
subtitle: "Atlit Common Tern & Little Tern ringing data - issue #5"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.width = 8, fig.height = 5)
library(readxl)
library(dplyr)
library(lubridate)
library(tidyr)
library(ggplot2)
library(knitr)
```

# Background

The paper is investigating what drives Little Tern breeding-success collapse
at Atlit, alongside the rise in Common Tern numbers. One candidate factor is
adult body condition (mass). This report builds and compares three versions
of a yearly, per-species mass metric from the raw ringing database
(`data_raw/ringing_data_raw.xlsx`, 66k+ individual records, local-only/not in
git - see `data_raw/README.md`):

1. **Spring (pre-breeding) mass** - the metric that reproduces Yosef's
   original `hirMASS`/`albMASS` columns (issue #5, task 1).
2. **Breeding-season mass** - restricted to June-July and to adults, as a
   proxy for actual Atlit breeders rather than passage migrants (issue #5,
   task 2).
3. **Confirmed-breeder spring mass** - a refinement of (1): spring-caught
   adults restricted further to individuals also encountered in June or
   July of the *same* year, i.e. more confident they stayed to breed at
   Atlit rather than just stopping over. Compared statistically against the
   full spring cohort.

This supersedes `scripts/02_reverse_engineer_mass.R` and
`scripts/03_breeding_mass.R`, consolidated here so all three metrics and
their comparisons are in one knittable report.

# Data loading

```{r load-data}
raw_dir <- "../data_raw"
if (!dir.exists(raw_dir)) raw_dir <- "data_raw"
stopifnot(dir.exists(raw_dir))

proc_dir <- "../data_processed"
if (!dir.exists(proc_dir)) proc_dir <- "data_processed"

out_dir <- "../output"
if (!dir.exists(out_dir)) out_dir <- "output"

ringing_path <- file.path(raw_dir, "ringing_data_raw.xlsx")
stopifnot(file.exists(ringing_path))

# The sheet has a 2-row merged header with duplicate leaf names (two "Date"
# columns, two "CR" columns), which readxl's auto name-repair mangles
# unpredictably. Skip both header rows and assign names explicitly, using
# the column mapping verified against data_raw/README.md. The leading "#"
# column (col 1 in the spreadsheet) is entirely empty and readxl drops it
# outright, so the real data is 47 columns, not 48.
col_names_47 <- c(
  "CR1","RingReplace","Rec","RingPrefix","RingNum","Species","Sex","Age","Wing","Weight",
  "Date","Time","Place","Lat","Lon","Ringer","Country","Remarks","Net","CR2","Subsp","Tail","Head",
  "Breeding","CP1","CP2","CP3","CP4","Colour","D","M","OrigDate","OrigCountry","OrigLat","OrigLon","OrigAge",
  "OldCR1","OldCR2","OldCR3","OldMetal1","OldMetal2","OilHead","OilWings","OilUpperP","OilUnderP","OilLegs","OilSum"
)
raw <- read_excel(ringing_path, sheet = "Data", skip = 2, col_names = col_names_47)

# EURING age codes present in this sheet: 1 = pullus (nestling), 3 = this
# calendar year juvenile (this season's fledgling); 4/5/6 and the two-digit
# known-exact-hatch-year codes (14, 22, 25, 28, 29, 35, 37...) are all
# adults (terns can live ~20 years, extending the scale for known-age
# birds - confirmed with Inbal).
juvenile_ages <- c(1, 3)

terns <- raw %>%
  filter(Species %in% c("STEALB", "STEHIR")) %>%
  mutate(
    year = year(Date),
    month = month(Date),
    ring_id = paste0(trimws(RingPrefix), RingNum),
    is_new = is.na(Rec) | trimws(Rec) == "",
    is_adult = !(Age %in% juvenile_ages)
  )

cat("Total tern records:", nrow(terns), "| distinct individuals (ring_id):", n_distinct(terns$ring_id), "\n")
```

**Caveat**: `ring_id` (`RingPrefix`+`RingNum`) identifies an individual by
its physical ring. `r sum(!is.na(raw$RingReplace))` records across the whole
sheet are flagged `RingReplace` (ring physically replaced) - for those
individuals, matching by `ring_id` across records breaks at the replacement
point. Not corrected for here; a source of (probably small) undercount in
the matching used in the "confirmed breeder" section below.

# Task 1: Spring (pre-breeding) mass

**Hypothesis** (confirmed in a prior session): `hirMASS`/`albMASS` = mean
`Weight` of new (non-recapture) individuals caught in spring (March-May),
per species per year.

```{r spring-mass}
spring_records <- terns %>%
  filter(is_new, !is.na(Weight), month %in% 3:5)

spring_mass <- spring_records %>%
  group_by(year, Species) %>%
  summarise(mean_weight = mean(Weight), sd_weight = sd(Weight), n = n(), .groups = "drop") %>%
  arrange(Species, year)

kable(spring_mass, digits = 2, caption = "Spring (Mar-May) mean weight, new individuals only, all ages")
```

## Validation against the known legacy values (2012-2020)

```{r spring-mass-validate}
legacy1 <- read.csv(file.path(raw_dir, "data1_2012-2020_legacy.csv"))
known <- legacy1 %>%
  select(year, hirMASS, albMASS) %>%
  pivot_longer(c(hirMASS, albMASS), names_to = "species_col", values_to = "reported_mass") %>%
  mutate(Species = ifelse(species_col == "hirMASS", "STEHIR", "STEALB"))

validation <- known %>%
  left_join(spring_mass, by = c("year", "Species")) %>%
  mutate(diff = mean_weight - reported_mass) %>%
  arrange(Species, year)

kable(validation %>% select(year, Species, reported_mass, computed = mean_weight, n, diff), digits = 2)

spring_cor <- cor(validation$reported_mass, validation$mean_weight, use = "complete.obs")
spring_mad <- mean(abs(validation$diff), na.rm = TRUE)
```

Correlation (reported vs. computed): **`r round(spring_cor, 4)`**. Mean
absolute difference: **`r round(spring_mad, 3)` g**. Confirms the spring
metric reproduces Yosef's original values almost exactly.

```{r spring-mass-plot}
ggplot(spring_mass, aes(year, mean_weight, color = Species)) +
  geom_line() + geom_point(aes(size = n)) +
  labs(title = "Spring (pre-breeding) mean mass by year", y = "Mean weight (g)", size = "n") +
  theme_minimal()
```

# Task 2: Breeding-season (June-July, adults only) mass

**Discovery** (not obvious from the task's wording, confirmed with Inbal
before implementing): a naive "new individuals caught in June-July" filter
is dominated by **this-year chicks/juveniles**, since chicks are ringed in
the nest/shortly after fledging throughout the breeding season - not by
adults. This must be filtered out or the metric mostly tracks fledgling
weight, not adult condition.

```{r age-month-evidence}
age_month_evidence <- terns %>%
  filter(is_new, month %in% 6:7) %>%
  count(Species, month, Age) %>%
  arrange(Species, month, desc(n)) %>%
  group_by(Species, month) %>%
  slice_max(n, n = 3) %>%
  ungroup()

kable(age_month_evidence, caption = "Top age codes by count, new individuals, Jun/Jul (age 1 = pullus, 3 = this-year juvenile - both excluded below)")
```

```{r breeding-mass}
breeding_records <- terns %>%
  filter(is_new, is_adult, !is.na(Weight), month %in% 6:7)

breeding_mass <- breeding_records %>%
  group_by(year, Species) %>%
  summarise(mean_weight = mean(Weight), sd_weight = sd(Weight), n = n(), .groups = "drop") %>%
  arrange(Species, year)

kable(breeding_mass, digits = 2, caption = "Breeding-season (Jun-Jul) mean weight, new ADULT individuals only")
```

## Comparison: spring vs. breeding-season mass

```{r spring-vs-breeding}
spring_vs_breeding <- spring_mass %>%
  rename(spring_mean = mean_weight, spring_n = n) %>%
  select(year, Species, spring_mean, spring_n) %>%
  full_join(
    breeding_mass %>% rename(breeding_mean = mean_weight, breeding_n = n) %>%
      select(year, Species, breeding_mean, breeding_n),
    by = c("year", "Species")
  ) %>%
  mutate(diff = breeding_mean - spring_mean) %>%
  arrange(Species, year)

kable(spring_vs_breeding, digits = 2)
```

### Year-level pattern

```{r spring-vs-breeding-stats, results = "asis"}
for (sp in c("STEHIR", "STEALB")) {
  sub <- spring_vs_breeding %>% filter(Species == sp, !is.na(spring_mean), !is.na(breeding_mean))
  r <- cor(sub$spring_mean, sub$breeding_mean)
  cat("**", sp, "**: correlation (spring vs. breeding) = ", round(r, 3),
      ", mean difference (breeding - spring) = ", round(mean(sub$diff), 2), " g\n\n", sep = "")
}
```

### Significance test (individual-level)

The year-level correlation/difference above summarises the *pattern*
across years, but doesn't test whether spring and breeding-season weights
actually differ. `spring_records` and `breeding_records` are built from
disjoint ringing events (a bird's first-ever ringing record can only be
"new" once - see `is_new` in the Data loading section), so these are
independent samples of individuals, suitable for a two-sample test:

```{r spring-vs-breeding-test, results = "asis"}
for (sp in c("STEHIR", "STEALB")) {
  spring_w <- spring_records %>% filter(Species == sp) %>% pull(Weight)
  breeding_w <- breeding_records %>% filter(Species == sp) %>% pull(Weight)
  tt <- t.test(spring_w, breeding_w)
  wt <- wilcox.test(spring_w, breeding_w)
  cat("**", sp, "**: spring n=", length(spring_w), ", mean=", round(mean(spring_w), 2),
      "g; breeding n=", length(breeding_w), ", mean=", round(mean(breeding_w), 2),
      "g; Welch t-test p=", signif(tt$p.value, 3),
      ", Wilcoxon p=", signif(wt$p.value, 3), "\n\n", sep = "")
}
```

**Year-controlled robustness check** - as in task 2b, a pooled comparison
across years can be confounded if spring/breeding records happen to
cluster in different years. This model holds year fixed while estimating
the breeding-vs-spring effect:

```{r spring-vs-breeding-test-year-controlled, results = "asis"}
for (sp in c("STEHIR", "STEALB")) {
  d <- bind_rows(
    spring_records %>% filter(Species == sp) %>% mutate(period = "spring"),
    breeding_records %>% filter(Species == sp) %>% mutate(period = "breeding")
  ) %>%
    mutate(year = factor(year), period = factor(period, levels = c("spring", "breeding")))
  m <- lm(Weight ~ period + year, data = d)
  co <- summary(m)$coefficients
  row <- co[grepl("^period", rownames(co)), , drop = FALSE]
  cat("**", sp, "** (year-controlled): breeding-vs-spring effect = ",
      round(row[1, "Estimate"], 2), " g (SE ", round(row[1, "Std. Error"], 2),
      "), p = ", signif(row[1, "Pr(>|t|)"], 3), "\n\n", sep = "")
}
```

```{r spring-vs-breeding-plot}
spring_vs_breeding %>%
  select(year, Species, spring_mean, breeding_mean) %>%
  pivot_longer(c(spring_mean, breeding_mean), names_to = "period", values_to = "mean_weight") %>%
  filter(!is.na(mean_weight)) %>%
  ggplot(aes(year, mean_weight, color = period)) +
  geom_line() + geom_point() +
  facet_wrap(~Species, scales = "free_y") +
  labs(title = "Spring vs. breeding-season mean mass", y = "Mean weight (g)") +
  theme_minimal()
```

Breeding-season adults weigh consistently less than spring arrivals -
plausibly the cost of incubation/chick-rearing.

# Task 2b: Confirmed-breeder spring mass

**Kept for reference/transparency - task 2c (any-year, below) is the
chosen definition for downstream analysis as of 2026-09-08.**

**Goal**: check whether Yosef's original spring mass parameter (task 1 -
mean weight of **all** spring-caught adults, new ringings only) differs
from the weight of the subset we're more confident actually bred at
Atlit that year. Approach: take **any** adult weighed in spring - a new
ringing *or* an in-hand recapture, doesn't matter which, just that it was
caught and weighed in March-May - and flag it as a confirmed breeder if
that same ring was also recorded (recapture *or* field resighting,
weighed or not) in June or July of the **same year**.

This deliberately does NOT filter to `is_new` (unlike task 1's spring
metric) - a recaptured adult weighed again in spring is just as valid a
spring-weight data point as a newly-ringed one for this comparison.

```{r confirmed-breeder-flag}
summer_presence <- terns %>%
  filter(month %in% 6:7) %>%
  distinct(ring_id, year) %>%
  mutate(seen_summer = TRUE)

spring_adults <- terns %>%
  filter(is_adult, !is.na(Weight), month %in% 3:5) %>%
  left_join(summer_presence, by = c("ring_id", "year")) %>%
  mutate(confirmed_breeder = coalesce(seen_summer, FALSE))

kable(spring_adults %>% count(Species, confirmed_breeder),
      caption = "Spring adults (new + recaptures): confirmed (seen again Jun/Jul same year) vs. not")

kable(spring_adults %>% count(Species, confirmed_breeder, is_new) %>%
        mutate(is_new = ifelse(is_new, "new ringing", "recapture")),
      caption = "Same breakdown, split by whether the spring record itself was a new ringing or a recapture")
```

**Caveat**: since recaptures are now included, the same individual can
appear more than once in `spring_adults` if it was caught and weighed in
spring in more than one year - a mild violation of independence in the
pooled tests below (not corrected for, on top of the existing
`RingReplace` caveat noted in Data loading).

```{r confirmed-breeder-summary}
confirmed_summary <- spring_adults %>%
  group_by(Species, confirmed_breeder) %>%
  summarise(mean_weight = mean(Weight), sd_weight = sd(Weight), n = n(), .groups = "drop")

kable(confirmed_summary, digits = 2, caption = "Spring weight: confirmed breeders vs. unconfirmed spring adults")
```

```{r confirmed-breeder-by-year}
confirmed_by_year <- spring_adults %>%
  group_by(year, Species, confirmed_breeder) %>%
  summarise(mean_weight = mean(Weight), n = n(), .groups = "drop")

kable(confirmed_by_year, digits = 2)
```

```{r confirmed-breeder-plot}
ggplot(confirmed_by_year, aes(year, mean_weight, color = confirmed_breeder, size = n)) +
  geom_point() + geom_line(aes(group = confirmed_breeder)) +
  facet_wrap(~Species, scales = "free_y") +
  labs(title = "Spring mass: confirmed breeders vs. unconfirmed, by year",
       y = "Mean weight (g)", color = "Confirmed\nbreeder") +
  theme_minimal()
```

## Significance tests

Two comparisons. Note "confirmed breeders" is no longer a strict subset of
task 1's "all spring individuals" cohort - task 1 is `is_new`-only (that's
how it reproduces Yosef's `hirMASS`/`albMASS`), while confirmed breeders
here can include recaptured adults too. So (a) compares confirmed breeders
against Yosef's original spring parameter as defined, and (b) is the
cleaner apples-to-apples split within the broader (new+recapture) spring
adult cohort built here:

**(a) Confirmed breeders vs. Yosef's original spring parameter** (task 1's
`is_new`-only cohort, as originally computed - checks whether the
breeder-confirmed subset differs from the parameter as Yosef defined it):

```{r test-vs-all-spring, results = "asis"}
for (sp in c("STEHIR", "STEALB")) {
  confirmed_w <- spring_adults %>% filter(Species == sp, confirmed_breeder) %>% pull(Weight)
  all_spring_w <- spring_records %>% filter(Species == sp) %>% pull(Weight)
  tt <- t.test(confirmed_w, all_spring_w)
  cat("**", sp, "**: confirmed n=", length(confirmed_w), ", mean=", round(mean(confirmed_w), 2),
      "g; all-spring n=", length(all_spring_w), ", mean=", round(mean(all_spring_w), 2),
      "g; Welch t-test p=", signif(tt$p.value, 3), "\n\n", sep = "")
}
```

**(b) Confirmed vs. unconfirmed spring adults** (non-overlapping groups -
the cleaner comparison, isolating "stayed" vs. "didn't stay" among adults):

```{r test-vs-unconfirmed, results = "asis"}
for (sp in c("STEHIR", "STEALB")) {
  confirmed_w <- spring_adults %>% filter(Species == sp, confirmed_breeder) %>% pull(Weight)
  unconfirmed_w <- spring_adults %>% filter(Species == sp, !confirmed_breeder) %>% pull(Weight)
  tt <- t.test(confirmed_w, unconfirmed_w)
  wt <- wilcox.test(confirmed_w, unconfirmed_w)
  cat("**", sp, "**: confirmed n=", length(confirmed_w), ", mean=", round(mean(confirmed_w), 2),
      "g; unconfirmed n=", length(unconfirmed_w), ", mean=", round(mean(unconfirmed_w), 2),
      "g; Welch t-test p=", signif(tt$p.value, 3),
      ", Wilcoxon p=", signif(wt$p.value, 3), "\n\n", sep = "")
}
```

**(c) Year-controlled robustness check.** Both groups' weights vary by
year, which could confound a naive pooled comparison if "confirmed"
records happen to cluster in heavier- or lighter-than-average years. This
model estimates the confirmed-breeder effect while holding year fixed:

```{r test-year-controlled, results = "asis"}
for (sp in c("STEHIR", "STEALB")) {
  d <- spring_adults %>% filter(Species == sp) %>% mutate(year = factor(year))
  m <- lm(Weight ~ confirmed_breeder + year, data = d)
  co <- summary(m)$coefficients
  row <- co[grepl("confirmed_breeder", rownames(co)), , drop = FALSE]
  cat("**", sp, "** (year-controlled): confirmed-breeder effect = ",
      round(row[1, "Estimate"], 2), " g (SE ", round(row[1, "Std. Error"], 2),
      "), p = ", signif(row[1, "Pr(>|t|)"], 3), "\n\n", sep = "")
}
```

## Interpretation

Read the three tests together per species rather than any single p-value:
(a) and (b) tell you the raw pooled-across-years difference; (c) tells you
whether that difference survives once year-to-year variation is accounted
for. Small per-year sample sizes (see the per-year table above) mean this
should be treated as an exploratory signal, not a final result, until
Yosef/Inbal review sample sizes per group per year.

# Task 2c: Confirmed-breeder spring mass - any-year site fidelity

**This is the chosen definition for downstream analysis** (decided
2026-09-08, after comparing both side by side - see the 2b vs. 2c
comparison table at the end of this section). Task 2b (same-year) is kept
above for reference/transparency, not deleted.

**Rationale** (Inbal): terns are typically site-faithful, returning to the
same colony year after year. Task 2b requires the summer sighting to fall
in the *same year* as the spring capture. This version relaxes that: a
bird recorded in June/July of **any** year (not necessarily the same year
as the spring weighing) is taken as evidence it's a colony-associated
breeder, and its spring weight - from whichever year it was caught in
spring - counts as a "confirmed breeder" data point. This recovers a
larger confirmed sample than the same-year version (roughly 2.5-3x), at
the cost of a looser link between the specific spring weight and evidence
of breeding *that particular* year (e.g. a bird could be flagged confirmed
based on a summer sighting a decade after or before the spring weight was
recorded).

```{r confirmed-breeder-anyyear-flag}
summer_ring_ids_anyyear <- terns %>%
  filter(month %in% 6:7) %>%
  distinct(ring_id) %>%
  pull(ring_id)

spring_adults_anyyear <- spring_adults %>%
  mutate(confirmed_breeder_anyyear = ring_id %in% summer_ring_ids_anyyear)

kable(spring_adults_anyyear %>% count(Species, confirmed_breeder_anyyear),
      caption = "Spring adults: confirmed via a summer sighting in ANY year vs. never recorded in summer")
```

```{r confirmed-breeder-anyyear-summary}
confirmed_summary_anyyear <- spring_adults_anyyear %>%
  group_by(Species, confirmed_breeder_anyyear) %>%
  summarise(mean_weight = mean(Weight), sd_weight = sd(Weight), n = n(), .groups = "drop")

kable(confirmed_summary_anyyear, digits = 2,
      caption = "Spring weight: any-year confirmed breeders vs. never recorded in summer")
```

```{r confirmed-breeder-anyyear-by-year}
confirmed_by_year_anyyear <- spring_adults_anyyear %>%
  group_by(year, Species, confirmed_breeder_anyyear) %>%
  summarise(mean_weight = mean(Weight), n = n(), .groups = "drop")

kable(confirmed_by_year_anyyear, digits = 2)
```

```{r confirmed-breeder-anyyear-plot}
ggplot(confirmed_by_year_anyyear, aes(year, mean_weight, color = confirmed_breeder_anyyear)) +
  geom_point() + geom_line(aes(group = confirmed_breeder_anyyear)) +
  facet_wrap(~Species, scales = "free_y") +
  labs(title = "Spring mass: any-year confirmed breeders vs. never recorded in summer",
       y = "Mean weight (g)", color = "Confirmed\nbreeder\n(any year)") +
  theme_minimal()
```

## Significance tests (any-year version)

Same three-test structure as task 2b, substituting the any-year flag:

```{r test-anyyear-vs-all-spring, results = "asis"}
for (sp in c("STEHIR", "STEALB")) {
  confirmed_w <- spring_adults_anyyear %>% filter(Species == sp, confirmed_breeder_anyyear) %>% pull(Weight)
  all_spring_w <- spring_records %>% filter(Species == sp) %>% pull(Weight)
  tt <- t.test(confirmed_w, all_spring_w)
  cat("**", sp, "**: confirmed (any-year) n=", length(confirmed_w), ", mean=", round(mean(confirmed_w), 2),
      "g; all-spring (Yosef's parameter) n=", length(all_spring_w), ", mean=", round(mean(all_spring_w), 2),
      "g; Welch t-test p=", signif(tt$p.value, 3), "\n\n", sep = "")
}
```

```{r test-anyyear-vs-unconfirmed, results = "asis"}
for (sp in c("STEHIR", "STEALB")) {
  confirmed_w <- spring_adults_anyyear %>% filter(Species == sp, confirmed_breeder_anyyear) %>% pull(Weight)
  unconfirmed_w <- spring_adults_anyyear %>% filter(Species == sp, !confirmed_breeder_anyyear) %>% pull(Weight)
  tt <- t.test(confirmed_w, unconfirmed_w)
  wt <- wilcox.test(confirmed_w, unconfirmed_w)
  cat("**", sp, "**: confirmed (any-year) n=", length(confirmed_w), ", mean=", round(mean(confirmed_w), 2),
      "g; never-seen-in-summer n=", length(unconfirmed_w), ", mean=", round(mean(unconfirmed_w), 2),
      "g; Welch t-test p=", signif(tt$p.value, 3),
      ", Wilcoxon p=", signif(wt$p.value, 3), "\n\n", sep = "")
}
```

```{r test-anyyear-year-controlled, results = "asis"}
for (sp in c("STEHIR", "STEALB")) {
  d <- spring_adults_anyyear %>% filter(Species == sp) %>% mutate(year = factor(year))
  m <- lm(Weight ~ confirmed_breeder_anyyear + year, data = d)
  co <- summary(m)$coefficients
  row <- co[grepl("confirmed_breeder_anyyear", rownames(co)), , drop = FALSE]
  cat("**", sp, "** (year-controlled): confirmed-breeder(any-year) effect = ",
      round(row[1, "Estimate"], 2), " g (SE ", round(row[1, "Std. Error"], 2),
      "), p = ", signif(row[1, "Pr(>|t|)"], 3), "\n\n", sep = "")
}
```

## Same-year (2b) vs. any-year (2c) comparison

```{r compare-2b-2c}
compare_2b_2c <- bind_rows(
  spring_adults %>% mutate(definition = "same-year (2b)") %>% rename(confirmed = confirmed_breeder),
  spring_adults_anyyear %>% mutate(definition = "any-year (2c)") %>% rename(confirmed = confirmed_breeder_anyyear)
) %>%
  filter(confirmed) %>%
  group_by(Species, definition) %>%
  summarise(mean_weight = mean(Weight), sd_weight = sd(Weight), n = n(), .groups = "drop")

kable(compare_2b_2c, digits = 2, caption = "Confirmed-breeder group only: same-year vs. any-year definition")
```

# Save outputs

```{r save-outputs}
write.csv(spring_mass, file.path(proc_dir, "spring_mass_by_year_species.csv"), row.names = FALSE)
write.csv(breeding_mass, file.path(proc_dir, "breeding_mass_by_year_species.csv"), row.names = FALSE)
write.csv(
  spring_adults_anyyear %>%
    select(ring_id, year, Species, Age, Weight, is_new, confirmed_breeder, confirmed_breeder_anyyear),
  file.path(proc_dir, "confirmed_breeder_spring_mass.csv"),
  row.names = FALSE
)

# ---- Update master_dataset_2010_2026.csv's hirMASS/albMASS columns -----
# Spring metric remains the single source for hirMASS/albMASS across all
# years (2011-2026), as established in the prior session - superseding
# data1.csv's 2012-2020-only values.
mass_wide <- spring_mass %>%
  mutate(col = ifelse(Species == "STEHIR", "hirMASS_new", "albMASS_new"),
         mean_weight = round(mean_weight, 1)) %>%
  select(year, col, mean_weight) %>%
  pivot_wider(names_from = col, values_from = mean_weight)

# ---- Add hirMASS_breed/albMASS_breed: confirmed-breeder spring mass ----
# Uses the chosen (task 2c, any-year site-fidelity) confirmed-breeder
# definition - mean spring Weight of adults also recorded in June/July of
# ANY year. Years with no confirmed breeders for a species are left NA.
breed_mass_wide <- spring_adults_anyyear %>%
  filter(confirmed_breeder_anyyear) %>%
  group_by(year, Species) %>%
  summarise(mean_weight = round(mean(Weight), 1), .groups = "drop") %>%
  mutate(col = ifelse(Species == "STEHIR", "hirMASS_breed", "albMASS_breed")) %>%
  select(year, col, mean_weight) %>%
  pivot_wider(names_from = col, values_from = mean_weight)

master_path <- file.path(proc_dir, "master_dataset_2010_2026.csv")
master <- read.csv(master_path) %>%
  select(-hirMASS, -albMASS, -any_of(c("hirMASS_breed", "albMASS_breed"))) %>%
  left_join(mass_wide, by = "year") %>%
  left_join(breed_mass_wide, by = "year") %>%
  rename(hirMASS = hirMASS_new, albMASS = albMASS_new)

write.csv(master, master_path, row.names = FALSE, na = "NA")
cat("Updated", master_path, "(added hirMASS_breed/albMASS_breed) and wrote 3 CSVs to", proc_dir, "\n")
```
