---
title: "Terns summer chick (juvenile) mass trends"
subtitle: "Issue #9 - follow-up to issue #8"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.width = 9, fig.height = 5)
library(readxl)
library(dplyr)
library(lubridate)
library(tidyr)
library(ggplot2)
library(knitr)
```

# Background

Same approach as #8, but for chicks: mass trend of this-year juveniles
(EURING age 3) caught in the colony during summer, **including recaptures
as well as new ringings** (previously #5 task 3 / #6 used new ringings
only). Window is **Jun-Aug** (not Jun-Jul like #8's adults) - matches the
juvenile window already established in #5 task 3 / #6, since juvenile
catches remain substantial into August (Common Tern age-3 catches:
Jun=106, Jul=1442, Aug=1049 - dropping August would discard a large,
non-trivial share of the data).

# Data loading

```{r load-data}
raw_dir <- "../data_raw"
if (!dir.exists(raw_dir)) raw_dir <- "data_raw"
stopifnot(dir.exists(raw_dir))

proc_dir <- "../data_processed"
if (!dir.exists(proc_dir)) proc_dir <- "data_processed"

out_dir <- "../output"
if (!dir.exists(out_dir)) out_dir <- "output"

ringing_path <- file.path(raw_dir, "ringing_data_raw.xlsx")
stopifnot(file.exists(ringing_path))

col_names_47 <- c(
  "CR1","RingReplace","Rec","RingPrefix","RingNum","Species","Sex","Age","Wing","Weight",
  "Date","Time","Place","Lat","Lon","Ringer","Country","Remarks","Net","CR2","Subsp","Tail","Head",
  "Breeding","CP1","CP2","CP3","CP4","Colour","D","M","OrigDate","OrigCountry","OrigLat","OrigLon","OrigAge",
  "OldCR1","OldCR2","OldCR3","OldMetal1","OldMetal2","OilHead","OilWings","OilUpperP","OilUnderP","OilLegs","OilSum"
)
raw <- read_excel(ringing_path, sheet = "Data", skip = 2, col_names = col_names_47)

terns <- raw %>%
  filter(Species %in% c("STEALB", "STEHIR")) %>%
  mutate(
    year = year(Date),
    month = month(Date),
    is_new = is.na(Rec) | trimws(Rec) == "",
    is_recap = trimws(Rec) == "R" & !is.na(Rec),
    is_capture = is_new | is_recap,  # excludes foreign-scheme codes (distant reports, not in-hand)
    species_label = ifelse(Species == "STEHIR", "Common Tern", "Little Tern")
  )
```

# Building the summer-chick group

```{r groups}
summer_chicks_raw <- terns %>%
  filter(is_capture, Age == 3, !is.na(Weight), month %in% 6:8) %>%
  mutate(group = "Summer chick")

n_before <- nrow(summer_chicks_raw)
summer_chicks <- summer_chicks_raw %>% filter(!is.na(Wing))
n_after <- nrow(summer_chicks)

cat("Jun-Aug age-3 records with Weight:", n_before, "- also with Wing:", n_after,
    sprintf("(%.1f%% retained)\n", 100 * n_after / n_before))
```

```{r coverage-table}
kable(
  summer_chicks %>% group_by(species_label) %>%
    summarise(n = n(), n_years = n_distinct(year),
              min_year = min(year), max_year = max(year), .groups = "drop"),
  caption = "Summer-chick weighed records by species"
)
```

```{r coverage-by-year}
kable(
  summer_chicks %>% count(species_label, year, name = "n") %>%
    pivot_wider(names_from = species_label, values_from = n, values_fill = 0),
  caption = "Sample size by year and species"
)
```

# Two normalization methods

Same approach as #6: **simple ratio** (`Weight/Wing`) and **SMI** (scaled
mass index, Peig & Green 2009 - log-log SMA regression). Both fit **per
species** (no group split needed here, since there's only one group per
species).

```{r normalize}
compute_smi <- function(df) {
  lx <- log(df$Wing)
  ly <- log(df$Weight)
  b <- sd(ly) / sd(lx)  # SMA slope
  wing_mean <- mean(df$Wing)
  df$smi <- df$Weight * (wing_mean / df$Wing) ^ b
  df$smi_b <- b
  df
}

summer_chicks <- summer_chicks %>%
  mutate(mass_ratio = Weight / Wing) %>%
  group_by(species_label) %>%
  group_modify(~ compute_smi(.x)) %>%
  ungroup()

kable(
  summer_chicks %>% distinct(species_label, smi_b),
  digits = 3, caption = "SMA slope (b) used for SMI, by species"
)
```

# Trend statistics, all three metrics

```{r trend-stats}
format_pval <- function(p) {
  ifelse(p < 0.001, formatC(p, format = "e", digits = 2), formatC(p, format = "f", digits = 3))
}

fit_trend <- function(df, response) {
  df %>%
    group_by(species_label) %>%
    group_modify(~ {
      m <- lm(reformulate("year", response = response), data = .x)
      s <- summary(m)
      tibble(
        n = nrow(.x),
        n_years = n_distinct(.x$year),
        slope_per_year = round(coef(m)["year"], 4),
        se = round(s$coefficients["year", "Std. Error"], 4),
        p_value = signif(s$coefficients["year", "Pr(>|t|)"], 3),
        r_squared = round(s$r.squared, 3)
      )
    }) %>%
    ungroup() %>%
    mutate(metric = response)
}

trend_stats <- bind_rows(
  fit_trend(summer_chicks, "Weight"),
  fit_trend(summer_chicks, "mass_ratio"),
  fit_trend(summer_chicks, "smi")
) %>%
  mutate(metric = recode(metric, Weight = "Raw mass (g)",
                          mass_ratio = "Mass/Wing ratio",
                          smi = "SMI (g, rescaled)"),
         metric = factor(metric, levels = c("Raw mass (g)", "Mass/Wing ratio", "SMI (g, rescaled)"))) %>%
  arrange(species_label, metric)

trend_stats_display <- trend_stats %>%
  mutate(significant = p_value < 0.05, p_value_fmt = format_pval(p_value)) %>%
  mutate(p_value_fmt = ifelse(significant, paste0("<strong>", p_value_fmt, "</strong>"), p_value_fmt)) %>%
  select(species_label, metric, n, n_years, slope_per_year, se, p_value = p_value_fmt, r_squared)

kable(trend_stats_display, escape = FALSE,
      caption = "value ~ year regression, by species/metric (significant rows, p&lt;0.05, in bold)")
```

# Visual comparison

Per-year mean shown as a point, with the **sample size (n) printed below
each point** - year-to-year coverage varies, and that should stay visible
rather than being smoothed over. Trend line fit on individual-level
records (not the per-year means), same convention as `03_mass_trends.Rmd`,
and only drawn where the year slope is significant (p<0.05).

```{r trends-plot, fig.height=7}
panel_levels <- c("Common Tern - Raw mass (g)", "Common Tern - Mass/Wing ratio", "Common Tern - SMI (g, rescaled)",
                   "Little Tern - Raw mass (g)", "Little Tern - Mass/Wing ratio", "Little Tern - SMI (g, rescaled)")

yearly_means <- summer_chicks %>%
  pivot_longer(cols = c(Weight, mass_ratio, smi), names_to = "metric", values_to = "value") %>%
  mutate(metric = recode(metric, Weight = "Raw mass (g)",
                          mass_ratio = "Mass/Wing ratio",
                          smi = "SMI (g, rescaled)"),
         panel_label = factor(paste(species_label, metric, sep = " - "), levels = panel_levels)) %>%
  group_by(species_label, metric, panel_label, year) %>%
  summarise(mean_value = mean(value), n = n(), .groups = "drop")

individual_long <- summer_chicks %>%
  pivot_longer(cols = c(Weight, mass_ratio, smi), names_to = "metric", values_to = "value") %>%
  mutate(metric = recode(metric, Weight = "Raw mass (g)",
                          mass_ratio = "Mass/Wing ratio",
                          smi = "SMI (g, rescaled)"),
         panel_label = factor(paste(species_label, metric, sep = " - "), levels = panel_levels))

sig_combos <- trend_stats %>% filter(p_value < 0.05) %>%
  mutate(metric = as.character(metric)) %>%
  distinct(species_label, metric)
individual_long_sig <- individual_long %>%
  mutate(metric = as.character(metric)) %>%
  semi_join(sig_combos, by = c("species_label", "metric"))

# n-count labels placed just below each panel's data range so they don't
# overlap the points/line, independent per panel (free_y scales).
label_pos <- yearly_means %>%
  group_by(panel_label) %>%
  mutate(y_label = min(mean_value) - 0.1 * (max(mean_value) - min(mean_value))) %>%
  ungroup()

ggplot(yearly_means, aes(year, mean_value, color = species_label)) +
  geom_point(size = 2) +
  geom_line(alpha = 0.3, linewidth = 0.4) +
  geom_smooth(data = individual_long_sig, aes(x = year, y = value), method = "lm", se = FALSE, linewidth = 0.8) +
  geom_text(data = label_pos, aes(y = y_label, label = n), size = 2.6, show.legend = FALSE) +
  facet_wrap(~panel_label, ncol = 3, scales = "free_y") +
  guides(color = "none") +
  labs(title = "Summer chick (Jun-Aug) mass trend by year and metric",
       subtitle = "Point = per-year mean; number below each point = sample size (n) that year. Trend line shown only where significant (p<0.05).",
       y = NULL, x = "Year") +
  theme_minimal()
```

# Save outputs

```{r save-outputs}
write.csv(summer_chicks %>% select(-smi_b),
          file.path(proc_dir, "summer_chick_mass_records.csv"), row.names = FALSE)
write.csv(trend_stats, file.path(proc_dir, "summer_chick_mass_trends_stats.csv"), row.names = FALSE)
cat("Wrote data_processed/summer_chick_mass_records.csv and summer_chick_mass_trends_stats.csv\n")
```
