---
title: "Terns within-season (summer) adult mass trajectory"
subtitle: "Issue #10 - follow-up to issue #8"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.width = 9, fig.height = 5)
library(readxl)
library(dplyr)
library(lubridate)
library(tidyr)
library(ggplot2)
library(knitr)
```

# Background

#5 task 4 asked whether adult mass changes over the course of the summer
(early vs. late season), but used the old spring-confirmed-breeder
population, which has no *weighed* within-season record after 2023 for
either species - so it couldn't say anything about recent years.

#8 established a better-covered population: adults caught **directly in
the colony during summer** (new ringings + recaptures, no breeder/non-
breeder split, since summer colony presence is itself the breeding
evidence) - data through 2026 for Common Tern. This script re-asks the
within-season question on that population, for **two window definitions**
run separately: **Jun-Jul** (matches #8) and **Jun-Aug** (matches the
chick window from #9, wider). For each window: is there a within-season
mass pattern **in general** (pooled across years), and does that pattern
**hold within each year on its own** (day-of-season x year interaction),
or does it vary?

# Data loading

```{r load-data}
raw_dir <- "../data_raw"
if (!dir.exists(raw_dir)) raw_dir <- "data_raw"
stopifnot(dir.exists(raw_dir))

proc_dir <- "../data_processed"
if (!dir.exists(proc_dir)) proc_dir <- "data_processed"

out_dir <- "../output"
if (!dir.exists(out_dir)) out_dir <- "output"

ringing_path <- file.path(raw_dir, "ringing_data_raw.xlsx")
stopifnot(file.exists(ringing_path))

col_names_47 <- c(
  "CR1","RingReplace","Rec","RingPrefix","RingNum","Species","Sex","Age","Wing","Weight",
  "Date","Time","Place","Lat","Lon","Ringer","Country","Remarks","Net","CR2","Subsp","Tail","Head",
  "Breeding","CP1","CP2","CP3","CP4","Colour","D","M","OrigDate","OrigCountry","OrigLat","OrigLon","OrigAge",
  "OldCR1","OldCR2","OldCR3","OldMetal1","OldMetal2","OilHead","OilWings","OilUpperP","OilUnderP","OilLegs","OilSum"
)
raw <- read_excel(ringing_path, sheet = "Data", skip = 2, col_names = col_names_47)

terns <- raw %>%
  filter(Species %in% c("STEALB", "STEHIR")) %>%
  mutate(
    year = year(Date),
    month = month(Date),
    date = as.Date(Date),
    is_new = is.na(Rec) | trimws(Rec) == "",
    is_recap = trimws(Rec) == "R" & !is.na(Rec),
    is_capture = is_new | is_recap,  # excludes foreign-scheme codes (distant reports, not in-hand)
    species_label = ifelse(Species == "STEHIR", "Common Tern", "Little Tern")
  )

juvenile_ages <- c(1, 3)

build_window <- function(months) {
  terns %>%
    filter(is_capture, !(Age %in% juvenile_ages), !is.na(Weight), month %in% months) %>%
    # Date comes in as POSIXct from readxl, so a naive Date - Date
    # subtraction returns a difftime in SECONDS, not days (bug caught in
    # #5 task 4) - force both sides to Date class first.
    mutate(day_of_season = as.numeric(date - as.Date(ymd(paste0(year, "-06-01")))))
}
```

# Analysis function

Reused identically for both window definitions below - pooled
`Weight ~ day_of_season` regression per species, plus a
`Weight ~ day_of_season * year` interaction test per species (does the
within-season slope vary from year to year), plus a per-year visual
(color-graded by year, following the convention from #5 task 4).

```{r analysis-fn}
format_pval <- function(p) {
  ifelse(p < 0.001, formatC(p, format = "e", digits = 2), formatC(p, format = "f", digits = 3))
}

run_within_season <- function(df, window_label) {
  pooled_stats <- df %>%
    group_by(species_label) %>%
    group_modify(~ {
      m <- lm(Weight ~ day_of_season, data = .x)
      s <- summary(m)
      tibble(
        n = nrow(.x),
        n_years = n_distinct(.x$year),
        slope_g_per_day = round(coef(m)["day_of_season"], 4),
        se = round(s$coefficients["day_of_season", "Std. Error"], 4),
        p_value = signif(s$coefficients["day_of_season", "Pr(>|t|)"], 3),
        r_squared = round(s$r.squared, 4)
      )
    }) %>%
    ungroup() %>%
    mutate(window = window_label)

  interaction_stats <- df %>%
    group_by(species_label) %>%
    filter(n_distinct(year) >= 3) %>%
    group_modify(~ {
      m <- lm(Weight ~ day_of_season * factor(year), data = .x)
      a <- anova(m)
      tibble(
        n = nrow(.x),
        n_years = n_distinct(.x$year),
        interaction_p = signif(a["day_of_season:factor(year)", "Pr(>F)"], 3)
      )
    }) %>%
    ungroup() %>%
    mutate(window = window_label)

  list(pooled = pooled_stats, interaction = interaction_stats, data = df %>% mutate(window = window_label))
}

# Year is a discrete quantity (16-17 distinct years), not a continuous one -
# scale_color_viridis_d (not _c) gives each year its own color from the
# viridis palette instead of implying a smooth gradient. The most recent
# year is additionally drawn as a thick black line on top, since it's the
# one most likely to get lost among 16 other overlapping lines/points but
# is usually the one of most immediate interest.
make_season_plot <- function(df, window_label, x_breaks, x_labels, show_points = TRUE) {
  df <- df %>% mutate(year_f = factor(year))
  latest_year <- max(df$year)
  df_latest <- df %>% filter(year == latest_year)
  # Excluded from the per-year colored-line layer below so the latest year
  # isn't drawn twice (once thin/colored, once thick/black) - the two lines
  # are numerically identical (same data, same lm fit) but overlapping them
  # made it look like two different trends at the line ends.
  df_hist <- df %>% filter(year != latest_year)

  p <- ggplot(df, aes(day_of_season, Weight, color = year_f))

  if (show_points) {
    p <- p + geom_jitter(width = 0.5, alpha = 0.35, size = 1)
  }

  p +
    geom_smooth(data = df_hist, aes(group = year_f), method = "lm", se = FALSE, linewidth = 0.5, alpha = 0.7) +
    geom_smooth(color = "grey30", method = "lm", se = FALSE, linewidth = 1, linetype = "dashed") +
    geom_smooth(data = df_latest, aes(x = day_of_season, y = Weight), inherit.aes = FALSE,
                method = "lm", se = FALSE, color = "black", linewidth = 1.6) +
    scale_color_viridis_d(name = "Year") +
    scale_x_continuous(breaks = x_breaks, labels = x_labels) +
    facet_wrap(~species_label, scales = "free_y") +
    labs(title = paste0("Adult mass within season (", window_label, ")",
                         if (show_points) "" else " - trend lines only"),
         subtitle = paste0(if (show_points) "Thin lines = per-year trend, " else "Per-year trend lines only (no raw points), ",
                            "grey dashed = pooled trend, thick black = most recent year (", latest_year, ")"),
         x = "Date within season", y = "Weight (g)") +
    theme_minimal()
}
```

# Window 1: Jun-Jul (no August)

```{r window-junjul}
w_junjul <- build_window(6:7)
kable(w_junjul %>% count(species_label, name = "n"), caption = "Jun-Jul adult weighed records by species")

res_junjul <- run_within_season(w_junjul, "Jun-Jul")

kable(res_junjul$pooled %>% select(-window), digits = 4,
      caption = "Pooled Weight ~ day_of_season, Jun-Jul (all years combined)")

kable(res_junjul$interaction %>% select(-window),
      caption = "day_of_season x year interaction test, Jun-Jul (does the within-season slope vary by year?)")
```

```{r window-junjul-plot, fig.height=5}
make_season_plot(w_junjul, "Jun-Jul", c(0, 30, 61), c("Jun 1", "Jul 1", "Jul 31"))
```

Same data, trend lines only (no raw points) - makes the most recent
year's line (thick black) easier to pick out against the rest:

```{r window-junjul-plot-lines, fig.height=5}
make_season_plot(w_junjul, "Jun-Jul", c(0, 30, 61), c("Jun 1", "Jul 1", "Jul 31"), show_points = FALSE)
```

# Window 2: Jun-Aug (including August)

```{r window-junaug}
w_junaug <- build_window(6:8)
kable(w_junaug %>% count(species_label, name = "n"), caption = "Jun-Aug adult weighed records by species")

res_junaug <- run_within_season(w_junaug, "Jun-Aug")

kable(res_junaug$pooled %>% select(-window), digits = 4,
      caption = "Pooled Weight ~ day_of_season, Jun-Aug (all years combined)")

kable(res_junaug$interaction %>% select(-window),
      caption = "day_of_season x year interaction test, Jun-Aug (does the within-season slope vary by year?)")
```

```{r window-junaug-plot, fig.height=5}
make_season_plot(w_junaug, "Jun-Aug", c(0, 30, 61, 91), c("Jun 1", "Jul 1", "Aug 1", "Aug 31"))
```

Same data, trend lines only (no raw points):

```{r window-junaug-plot-lines, fig.height=5}
make_season_plot(w_junaug, "Jun-Aug", c(0, 30, 61, 91), c("Jun 1", "Jul 1", "Aug 1", "Aug 31"), show_points = FALSE)
```

# Per-year slopes and recent-vs-historical test (Jun-Jul only)

August can include passage migrants as well as the local breeding
population, so this follow-up restricts to the Jun-Jul window only.

## Per-year within-season slope

Same Jun-Jul population as above, but fit separately per year instead of
pooled, so each year's own within-season trend is visible on its own.
Years with fewer than 5 weighed records are dropped (too few points to
fit a meaningful line) - in practice this only drops 2025 for Common
Tern, which has zero Jun-Jul records that year.

```{r peryear-slopes}
peryear_junjul <- w_junjul %>%
  group_by(species_label, year) %>%
  filter(n() >= 5) %>%
  group_modify(~ {
    m <- lm(Weight ~ day_of_season, data = .x)
    s <- summary(m)
    tibble(
      n = nrow(.x),
      slope_g_per_day = round(coef(m)["day_of_season"], 4),
      se = round(s$coefficients["day_of_season", "Std. Error"], 4),
      p_value = signif(s$coefficients["day_of_season", "Pr(>|t|)"], 3)
    )
  }) %>%
  ungroup()

kable(peryear_junjul, digits = 4,
      caption = "Per-year Weight ~ day_of_season slope, Jun-Jul (years with n >= 5 only)")
```

```{r peryear-signs}
kable(peryear_junjul %>%
        group_by(species_label) %>%
        summarise(n_years = n(), n_negative = sum(slope_g_per_day < 0),
                  n_positive = sum(slope_g_per_day > 0), .groups = "drop"),
      caption = "Sign of the per-year slope, Jun-Jul")
```

**Common Tern: 11 of 16 years positive** (mass tends to increase over
Jun-Jul most years). **Little Tern: 11 of 17 years negative** (mass tends
to decrease over Jun-Jul most years) - the opposite direction from
Common Tern, matching the pooled Jun-Jul result above. Individual years
are mostly not significant on their own (Common Tern: 2016, 2020, 2021,
2024 are; Little Tern: none are) - year-to-year slope estimates are
noisy, which is exactly why a pooled/interaction test (rather than
eyeballing single-year lines) is needed to say anything about whether a
given year, or the last few years, are actually different from the rest.

## Is the most recent year (or last few years) different from the rest?

**Method and why:** to test whether one group of years has a different
within-season slope than the rest, the raw records are pooled and fit
with `Weight ~ day_of_season * group`, where `group` is a 2-level factor
(recent vs. everyone else). The `day_of_season:group` interaction
coefficient is literally the difference in slope between the two groups,
with its own standard error and p-value. This is preferred over treating
the per-year slopes above as a small sample and asking whether the most
recent one is an outlier in that sample (e.g. a z-score against the
spread of slopes) - that approach would treat each year's slope as known
exactly, when in fact most individual years are too noisy to trust on
their own (see the p-value column above). Fitting on the pooled raw data
instead lets years with more/cleaner data count for more, and is the
same logic as the `day_of_season x year` interaction test used earlier
in this script, just with `year` collapsed into two groups instead of
kept as 17 separate levels.

Two group definitions are tested: the single most recent year (2026) vs.
all other years, and the last 3 years *with data* vs. earlier years
("with data" rather than last 3 calendar years, since Common Tern has
zero Jun-Jul records in 2025 - a calendar-year window would silently
collapse that comparison to just 2 recent years for that species).

```{r recent-vs-historical}
test_recent_vs_rest <- function(df, group_years, group_label) {
  d <- df %>%
    mutate(group = factor(ifelse(year %in% group_years, "recent", "other"),
                           levels = c("other", "recent")))
  m <- lm(Weight ~ day_of_season * group, data = d)
  s <- summary(m)
  slope_other <- coef(m)["day_of_season"]
  slope_diff <- coef(m)["day_of_season:grouprecent"]
  tibble(
    comparison = group_label,
    n_recent = sum(d$group == "recent"),
    n_other = sum(d$group == "other"),
    slope_other = round(slope_other, 4),
    slope_recent = round(slope_other + slope_diff, 4),
    slope_diff = round(slope_diff, 4),
    p_value = signif(s$coefficients["day_of_season:grouprecent", "Pr(>|t|)"], 3)
  )
}

recent_tests <- bind_rows(
  w_junjul %>% filter(species_label == "Common Tern") %>%
    test_recent_vs_rest(2026, "Common Tern: 2026 vs. all other years"),
  w_junjul %>% filter(species_label == "Common Tern") %>%
    test_recent_vs_rest(c(2023, 2024, 2026), "Common Tern: last 3 yrs w/ data (2023, 2024, 2026) vs. earlier"),
  w_junjul %>% filter(species_label == "Little Tern") %>%
    test_recent_vs_rest(2026, "Little Tern: 2026 vs. all other years"),
  w_junjul %>% filter(species_label == "Little Tern") %>%
    test_recent_vs_rest(c(2024, 2025, 2026), "Little Tern: last 3 yrs w/ data (2024, 2025, 2026) vs. earlier")
)

kable(recent_tests, digits = 4,
      caption = "Recent-years vs. historical within-season slope, Jun-Jul (day_of_season x group interaction test)")
```

**No significant difference for either species, under either definition
of "recent"** (all four interaction p-values > 0.15). Despite 2026 (and
the last few years) visually looking flat/down for Common Tern against a
pooled upward historical trend, that gap is not statistically
distinguishable from the year-to-year noise already present in the
historical data - individual "ordinary" years swing from -0.29 to +0.50
g/day (see per-year table above). This is a genuine null result rather
than an underpowered one in an absolute sense - the "other years" side of
each comparison has hundreds of records - but with only n=82 (Common
Tern) / n=23 (Little Tern) on the "recent" side, a modest true shift
could still be missed; read this as "no evidence of a recent shift," not
"definitively no shift."

# Interpretation

A significant pooled slope (first table in each window section) means
mass systematically changes over the course of that window, averaged
across all years. A significant `day_of_season x year` interaction means
that pattern isn't consistent from year to year - some years may show a
strong trend and others none, or even the opposite direction - so the
pooled slope should be read as an average tendency, not something that
holds every single year. A non-significant interaction means the pooled
slope is a reasonably good summary of what happens within any given
year's season on its own. Read both together with the per-year plot,
which makes any single outlier year visually obvious (color-graded by
year - a year that stands out from the rest is visible regardless of
whether it's an early or recent year).

# Save outputs

```{r save-outputs}
all_pooled <- bind_rows(res_junjul$pooled, res_junaug$pooled)
all_interaction <- bind_rows(res_junjul$interaction, res_junaug$interaction)

write.csv(all_pooled, file.path(proc_dir, "summer_adult_within_season_pooled_stats.csv"), row.names = FALSE)
write.csv(all_interaction, file.path(proc_dir, "summer_adult_within_season_interaction_stats.csv"), row.names = FALSE)
write.csv(peryear_junjul, file.path(proc_dir, "summer_adult_within_season_junjul_peryear_slopes.csv"), row.names = FALSE)
write.csv(recent_tests, file.path(proc_dir, "summer_adult_within_season_junjul_recent_vs_historical.csv"), row.names = FALSE)
cat("Wrote data_processed/summer_adult_within_season_pooled_stats.csv, summer_adult_within_season_interaction_stats.csv, summer_adult_within_season_junjul_peryear_slopes.csv and summer_adult_within_season_junjul_recent_vs_historical.csv\n")
```
